### 2022-06-28 preprint.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@1d38a4d0137defc303e636439c8f51544298a89c 🚀

In this section, we illustrate the embedding of a collection of
bipartite networks collected by Hadfield et al. (2014), using truncated
Singular Value Decomposition (t-SVD) and RDPG (see Strydom et al., 2022
for the full details). Briefly, an RDPG decomposes a network into two
subspaces (left and right), which are matrices that when multiplied give
an approximation of the original network. The code to reproduce this
example is available as supplementary material (note, for the sake of
comparison, that Strydom, Catchen, et al., 2021 have an example using
embedding through PCA followed by prediction using a deep neural network
on the same dataset). The resulting (binary) metaweb \(\mathcal{M}\) has
2131 interactions between 206 parasites and 121 hosts, and its adjacency
matrix has full rank (\emph{i.e.} it represents a space with 121
dimensions). All analyses were done using Julia (Bezanson et al., 2017)
version 1.7.2, \emph{Makie.jl} (Danisch \& Krumbiegel, 2021), and
\emph{EcologicalNetworks.jl} (Poisot et al., 2019).

@@

The embedding of the metaweb holds several pieces of information
(fig.~\ref{fig:illustration}). In panel \textbf{A}, we show that the
\(L_2\) loss (\emph{i.e.} the sum of squared errors) between the
empirical and reconstructed metaweb decreases when number of dimensions
(rank) of the subspace increases, with a point of inflection around 25
dimensions. As discussed by Runghen et al. (2021), there is often a
trade-off between the number of dimensions to use (more dimensions are
more computationally demanding) and the quality of the representation.
In this instance, accepting \(L_2 = 500\) as an approximation of the
network means that the error for every position in the metaweb is
\(\approx \left(500/(206\times 121)\right)^{1/2}\). In \textbf{B}, we
show the positions of hosts and parasites on the first two dimensions of
the left and right subspaces. Note that these values largely skew
negative, because the first dimensions capture the coarse structure of
the network: most pairs of species do not interact, and therefore have
negative values. In \textbf{C}, we show the predicted weight
(\emph{i.e.} the result of the multiplication of the RDGP subspaces at a
rank of 25) as a function of whether the interactions are observed,
not-observed, or unknown due to lack of co-occurence. This reveals that
the obserbed interactions have higher predicted weights, although there
is some overlap; the usual approach to identify possible interactions
based on this information would be a thresholding analysis, which is
outside the scope of this manuscript (and is done in the papers cited in
this illustration). Note that the values are not bound to the unit
interval, which emphasizes the need for either scaling or clamping
(although thresholding analyses are insensitive to this choice).
Finally, in \textbf{D}, we show that the embedding, as it captures
structural information about the network, holds ecological information;
indeed, the position of the parasite on the first dimension of the left
sub-space is a linear predictor of its number of hosts.

@@

The first open research problem is the taxonomic and spatial limit of
the metaweb to embed and transfer. If the initial metaweb is too narrow
in scope, notably from a taxonomic point of view, the chances of finding
another area with enough related species (through phylogenetic
relatedness or similarity of functional traits) to make a reliable
inference decreases; this would likely be indicated by large confidence
intervals during estimation of the values in the low-rank space, meaning
that the representation of the original graph is difficult to transfer
to the new problem. In addition, other problems can arise due to
non-overlapping trait distributions in the known and unknown species.
Alternatively a metaweb is too large (taxonomically), then the resulting
embeddings would need to represent interactions between taxonomic groups
that are not present in the new location. This would lead to a much
higher variance in the starting dataset, and to under-dispersal in the
target dataset, resulting in the potential under or over estimation of
the strength of new predicted interactions. The lack of well documented
metawebs is currently preventing the development of more concrete
guidelines. The question of phylogenetic relatedness and dispersal is
notably true if the metaweb is assembled in an area with mostly endemic
species (\emph{e.g.} a system that has undergone recent radiation and
might not have an analogous system with which to draw knowledge from),
and as with every predictive algorithm, there is room for the
application of our best ecological judgement. Because this problem
relates to dispersal of species in the geographic or phylogenetic space,
it can certainly be approached through assessing the performance of
embedding transfer in simulated starting/target species pools.

The second series of problems relate to determining which area should be
used to infer the new metaweb in, as this determines the species pool
that must be used. Metawebs can be constructed by assigning interactions
in a list of species within geographic boundaries. The upside of this
approach is that information at the country level is likely to be
required for biodiversity assessments, as countries set goals at the
national level (Buxton et al., 2021), and as quantitative instruments
are designed to work at these scales (Turak et al., 2017); specific
strategies are often enacted at smaller scales, nested within a specific
country (Ray et al., 2021). But there is no guarantee that these
boundaries are meaningful. In fact, we do not have a satisfying answer
to the question of ``where does a food web stop?''; the most promising
solutions involve examining the spatial consistency of network area
relationships (Fortin et al., 2021; see \emph{e.g.} Galiana et al.,
2018, 2019, 2021), which is impossible in the absence of enough
information about the network itself. This suggests that inferred
metawebs should be further downscaled to allow for \emph{a posteriori}
analyses. The methodology for metaweb downscaling is currently limited,
and it is likely that the sustained effort to characterize the spatial
dependency of food web structure will lead to more prescriptive
guidelines about the need for prediction downscaling.

The final family of problems relates less to ecological methods than to
the praxis of ecological research. Operating under the context of
national divisions, in large parts of the world, reflects nothing more
than the legacy of settler colonialism, which drives a disparity in
available ecological data. Applying any embedding to biased data does
not debias them, but instead embeds these very same biases, propagating
them to the machine learning models using embeddings tomake predictions.
Indeed, the use of ecological data is not an apolitical act (Nost \&
Goldstein, 2021), as data infrastructures tend to be designed to answer
questions within national boundaries (therefore placing contingencies on
what is available to be embedded), and their use often draws upon and
reinforces territorial statecraft. As per Machen \& Nost (2021), this is
particularly true when the output of ``algorithmic thinking''
(\emph{e.g.} relying on machine learning to generate knowledge) can be
re-used for governance (\emph{e.g.} enacting conservation decisions at
the national scale). As information on species interaction networks
structure is increasingly leveraged as a tool to guide conservation
actions (see \emph{e.g.} recent discussions for food-web based
conservation; Eero et al., 2021; Naman et al., 2022; Stier et al.,
2017), the need to appraise and correct biases that are unwittingly
propagated to algorithms when embedded from the original data is
paramount. Predictive approaches deployed at the continental scale, no
matter their intent, originate in the framework that contributed to the
ongoing biodiversity crisis (Adam, 2014) and reinforced environmental
injustice (Choudry, 2013; Domínguez \& Luoma, 2020). Particularly on
Turtle Island and other territories that were traditionally stewarded by
Indigenous people, these approaches should be replaced (or at least
guided and framed) by Indigenous principles of land management (Eichhorn
et al., 2019; No'kmaq et al., 2021), as part of an
``algorithm-in-the-loop'' approach. Human-algorithm interactions are
notoriously difficult and can yield adverse effect (Green \& Chen, 2019;
Stevenson \& Doleac, 2021), suggesting the need to systematically study
them for the specific purpose of biodiversity governance, as well as to
improve the algorithmic literacy of decision makers. As we see
artificial intelligence/machine learning being increasingly mobilized to
generate knowledge that is lacking for conservation decisions
(\emph{e.g.} Lamba et al., 2019; Mosebo Fernandes et al., 2020) and
drive policy decisions (Weiskopf et al., 2022), our discussion of these
tools need to go beyond the technical and statistical, and into the
governance consequences they can have.

@@

@@

>>>>

In this section, we illustrate the embedding of a collection of
bipartite networks collected by Hadfield et al. (2014), using t-SVD and
RDPG (see Strydom et al., 2022 for the full details). Briefly, an RDPG
decomposes a network into two subspaces (left and right), which are
matrices that when multiplied give an approximation of the original
network. The code to reproduce this example is available as
supplementary material (note, for the sake of comparison, that Strydom,
Catchen, et al., 2021 have an example using embedding through PCA
followed by prediction using a deep neural network on the same dataset).
The resulting (binary) metaweb \(\mathcal{M}\) has 2131 interactions
between 206 parasites and 121 hosts, and its adjacency matrix has full
rank (\emph{i.e.} it represents a space with 121 dimensions). All
analyses were done using Julia (Bezanson et al., 2017) version 1.7.2,
\emph{Makie.jl} (Danisch \& Krumbiegel, 2021), and
\emph{EcologicalNetworks.jl} (Poisot et al., 2019).

@@

The embedding of the metaweb holds several pieces of information
(fig.~\ref{fig:illustration}). In panel \textbf{A}, we show that the
\(L_2\) loss (\emph{i.e.} the sum of squared errors) between the
empirical and reconstructed metaweb decreases when the number of
dimensions (rank) of the subspace increases, with an inflection point
around 25 dimensions. As discussed by Runghen et al. (2021), there is
often a trade-off between the number of dimensions to use (more
dimensions are more computationally demanding) and the quality of the
representation. In this instance, accepting \(L_2 = 500\) as an
approximation of the network means that the error for every position in
the metaweb is \(\approx \left(500/(206\times 121)\right)^{1/2}\). In
fig.~\ref{fig:illustration}, panel \textbf{B}, we show the positions of
hosts and parasites on the first two dimensions of the left and right
subspaces. Note that these values largely skew negative, because the
first dimensions capture the coarse structure of the network: most pairs
of species do not interact, and therefore have negative values. In
fig.~\ref{fig:illustration}, panel \textbf{C}, we show the predicted
weight (\emph{i.e.} the result of the multiplication of the RDGP
subspaces at a rank of 25) as a function of whether the interactions are
observed, not-observed, or unknown due to lack of co-occurrence. This
reveals that the observed interactions have higher predicted weights,
although there is some overlap; the usual approach to identify potential
interactions based on this information would be a thresholding analysis,
which is outside the scope of this manuscript (and is done in the papers
cited in this illustration). Note that the values are not bound to the
unit interval, which emphasizes the need for either scaling or clamping
(although thresholding analyses are insensitive to this choice).
Finally, in fig.~\ref{fig:illustration}, panel \textbf{D}, we show that
the embedding, as it captures structural information about the network,
holds ecological information; indeed, the position of the parasite on
the first dimension of the left sub-space is a linear predictor of its
number of hosts.

@@

The first open research problem is the taxonomic and spatial limit of
the metaweb to embed and transfer. If the initial metaweb is too narrow
in scope, notably from a taxonomic point of view, the chances of finding
another area with enough related species (through phylogenetic
relatedness or similarity of functional traits) to make a reliable
inference decreases; this would likely be indicated by large confidence
intervals during estimation of the values in the low-rank space, meaning
that the representation of the original graph is difficult to transfer
to the new problem. Alternatively, if the initial metaweb is too large
(taxonomically), then the resulting embeddings would need to represent
interactions between taxonomic groups that are not present in the new
location. This would lead to a much higher variance in the starting
dataset, and to under-dispersion in the target dataset, resulting in the
potential under or over estimation of the strength of new predicted
interactions. The lack of well documented metawebs is currently
preventing the development of more concrete guidelines. The question of
phylogenetic relatedness and distribution is notably relevant if the
metaweb is assembled in an area with mostly endemic species (\emph{e.g.}
a system that has undergone recent radiation or that has remained in
isolation for a long period of time might not have an analogous system
with which to draw knowledge from), and as with every predictive
algorithm, there is room for the application of our best ecological
judgement. Because this problem relates to distribution of species in
the geographic or phylogenetic space, it can certainly be approached
through assessing the performance of embedding transfer in simulated
starting/target species pools.

The second series of problems relate to determining which area should be
used to infer the new metaweb in, as this determines the species pool
that must be used. Metawebs can be constructed by assigning interactions
in a list of species within geographic boundaries. The upside of this
approach is that information at the country level is likely to be
required for biodiversity assessments, as countries set conservation
goals at the national level (Buxton et al., 2021), and as quantitative
instruments are designed to work at these scales (Turak et al., 2017);
specific strategies are often enacted at smaller scales, nested within a
specific country (Ray et al., 2021). But there is no guarantee that
these boundaries are meaningful. In fact, we do not have a satisfying
answer to the question of ``where does a food web stop?''; the most
promising solutions involve examining the spatial consistency of network
area relationships (Fortin et al., 2021; see \emph{e.g.} Galiana et al.,
2018, 2019, 2021), which is impossible in the absence of enough
information about the network itself. This suggests that inferred
metawebs should be further downscaled to allow for \emph{a posteriori}
analyses. The methodology for metaweb downscaling is currently limited,
and it is likely that the sustained effort to characterize the spatial
dependency of food web structure will lead to more prescriptive
guidelines about the need for prediction downscaling.

The final family of problems relates less to ecological methods than to
the praxis of ecological research. Operating under the context of
national divisions, in large parts of the world, reflects nothing more
than the legacy of settler colonialism, which drives a disparity in
available ecological data. Applying any embedding to biased data does
not debias them, but instead embeds these very same biases, propagating
them to the machine learning models using embeddings to make
predictions. Indeed, the use of ecological data is not an apolitical act
(Nost \& Goldstein, 2021), as data infrastructures tend to be designed
to answer questions within national boundaries (therefore placing
contingencies on what is available to be embedded), and their use often
draws upon and reinforces territorial statecraft. As per Machen \& Nost
(2021), this is particularly true when the output of ``algorithmic
thinking'' (\emph{e.g.} relying on machine learning to generate
knowledge) can be re-used for governance (\emph{e.g.} enacting
conservation decisions at the national scale). As information on species
interaction networks structure is increasingly leveraged as a tool to
guide conservation actions (see \emph{e.g.} recent discussions for
food-web based conservation; Eero et al., 2021; Naman et al., 2022;
Stier et al., 2017), the need to appraise and correct biases that are
unwittingly propagated to algorithms when embedded from the original
data is paramount. Predictive approaches deployed at the continental
scale, no matter their intent, originate in the framework that
contributed to the ongoing biodiversity crisis (Adam, 2014) and
reinforced environmental injustice (Choudry, 2013; Domínguez \& Luoma,
2020). Particularly on Turtle Island and other territories that were
traditionally stewarded by Indigenous people, these approaches should be
replaced (or at least guided and framed) by Indigenous principles of
land management (Eichhorn et al., 2019; No'kmaq et al., 2021), as part
of an ``algorithm-in-the-loop'' approach. Human-algorithm interactions
are notoriously difficult and can yield adverse effect (Green \& Chen,
2019; Stevenson \& Doleac, 2021), suggesting the need to systematically
study them for the specific purpose of biodiversity governance, as well
as to improve the algorithmic literacy of decision makers. As we see
artificial intelligence/machine learning being increasingly mobilized to
generate knowledge that is lacking for conservation decisions
(\emph{e.g.} Lamba et al., 2019; Mosebo Fernandes et al., 2020) and
drive policy decisions (Weiskopf et al., 2022), our discussion of these
tools need to go beyond the technical and statistical, and into the
governance consequences they can have.

@@

\leavevmode\hypertarget{ref-Chami2022MacLea}{}%
Chami, I., Abu-El-Haija, S., Perozzi, B., Ré, C., \& Murphy, K. (2022).
Machine Learning on Graphs: A Model and Comprehensive Taxonomy.
\emph{Journal of Machine Learning Research}, \emph{23}(89), 1--64.
\url{http://jmlr.org/papers/v23/20-852.html}

@@

\leavevmode\hypertarget{ref-Gibb2021DatPro}{}%
Gibb, R., Albery, G. F., Becker, D. J., Brierley, L., Connor, R.,
Dallas, T. A., Eskew, E. A., Farrell, M. J., Rasmussen, A. L., Ryan, S.
J., Sweeny, A., Carlson, C. J., \& Poisot, T. (2021). Data
Proliferation, Reconciliation, and Synthesis in Viral Ecology.
\emph{BioScience}, \emph{71}(11), 1148--1156.
\url{https://doi.org/10.1093/biosci/biab080}
